Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-01-19 LaTeX/Turnover_plos.tex
PloS style edits complete. Sally edits to Appendix and References

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex ratio bias, and transitions between sex-determination systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institue, Department of Genetics, Evolution and Environment, University Collect London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Sex-determination systems are remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determination systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
(2) We also consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition) or meiosis (i.e., non-Mendelian segregation); selective processes that typically occur in one sex or the other. 
We find that associations with haploid selected loci can drive transitions between sex determination systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determination systems are likely to be labile and draw novel connections with sex ratio evolution

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
%Therefore, our models radically alter our understanding of when and why new sex determination system can evolve. 
%Therefore, our models cause us to update the prevailing understanding about selective forces that can drive transitions between sex determination systems.

@@

\externaldocument{Turnover_plos.tex}
\externaldocument{S1_Appendix.tex}
%\externaldocument{S2_Appendix.tex}
\externaldocument{S3_Appendix.tex}

@@

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles (Figure \ref{fig:model_outline}). 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, we can interpret $m$ as an environmental sex determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
%We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which assumes a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci. %(and the \textbf{M} genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Figure \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bears allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic, $\lambda^2+ b \lambda + c =0$ \textcolor{blue}{(see Appendix for a discussion of other roots - or Sally's proof!)}.
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\chi_{mA} + \chi_{ma})$ and $c = (\lambda_{mA}-\chi_{mA}) (\lambda_{ma}-\chi_{ma}) -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i$ without accounting for loss due to recombination, and $\chi_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\chi_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

>>>>

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figs should be uploaded separately from your manuscript file. 
% - Figs generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figs containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% Bold the 'Fig #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex ratio bias, and transitions between sex-determination systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 312 words
\section*{Abstract}
Sex-determination systems are remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determination systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
(2) We also consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition) or meiosis (i.e., non-Mendelian segregation); selective processes that typically occur in one sex or the other. 
We find that associations with haploid selected loci can drive transitions between sex determination systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determination systems are likely to be labile and draw novel connections with sex ratio evolution

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Accumulating evidence indicates that systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex determination systems? 
Here, we use models to derive conditions under which new sex determination systems spread. 
Prevailing views suggest that new sex determination systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which both causes sex ratio biases and occurs differently in males and females. 
Surprisingly, we find that neither force (sex ratio selection nor associations with genes that have sex-specific effects) dominates the spread of new sex determination systems alone. 
Even more unexpectedly, we find that new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females. 
Therefore, our models predict loci in unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex determination systems.

@@

\externaldocument{Turnover_plos.tex}
\externaldocument{S1_Appendix.tex}
\externaldocument{S2_Appendix.tex}
\externaldocument{S3_Appendix.tex}

@@

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles (Fig \ref{fig:model_outline}). 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, we can interpret $m$ as an environmental sex determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
%We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in \nameref{app:recurs}. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which assumes a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci. %(and the \textbf{M} genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bears allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic, $\lambda^2+ b \lambda + c =0$ \textcolor{blue}{(see \nameref{app:inv_cond} for a discussion of other roots - or Sally's proof!)}.
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\chi_{mA} + \chi_{ma})$ and $c = (\lambda_{mA}-\chi_{mA}) (\lambda_{ma}-\chi_{ma}) -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i$ without accounting for loss due to recombination, and $\chi_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\chi_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

\begin{table}[!ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{ {\bf Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   $m$ is a neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{YA} = {\left( 2 \zeta \right)}^{-1} \left[\hat{p}_X^\female w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{Ya} = {\left( 2 \zeta \right)}^{-1} \left[(1-\hat{p}_X^\female) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{YA} = R {\left( 2 \zeta \right)}^{-1} \left[ (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/  \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)   $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Ya} = R {\left( 2 \zeta \right)}^{-1} \left[   \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha_\Delta^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)  $\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  $m$ is a neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{WA} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}(1+\alpha_{\Delta}^{\female})\right]/ \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{Wa} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha_{\Delta}^{\female})\right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{WA} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1+\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right)  $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Wa} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [1ex]
  \hline \hline 
   \end{tabular}
   \begin{flushleft} 
$\bar{p}^{\male}=(1-\hat{q})\hat{p}_X^\male + \hat{q}\hat{p}_Y^\male$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
%      \item $R$ is the probability of recombination between loci \textbf{A} and \textbf{M}.
$\zeta$ is the zygotic sex ratio (fraction male).
$\bar{w}^{\Hermaphrodite}_{D}$ is the mean fitness of diploids of sex $\Hermaphrodite$, see Table \ref{tab:meanfitnesses}.
$\bar{w}_H^\Hermaphrodite$ is the mean fitness of haploids from sex $\Hermaphrodite$, see Table \ref{tab:meanfitnesses}.
\end{flushleft}
  \label{tab:haplotype_growth}
\end{table}
